---
title: "Project part Mark"
author: "Hanxi Chen"
subtitle: :)
date: today
format:
  pdf:
    documentclass: article
    number-sections: false
    fig-align: center
    keep-tex: false  # keeps the intermediate .tex file
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    html-math-method: mathjax
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r, warning=FALSE}
library(tidyverse)
library(car)
library(dplyr)
library(survival)
library(survminer)
library(zoo) # na.locf

library(janitor)
library(skimr)

library(tableone)

library(tableone)
library(gtsummary)
```


## Import and Cleaning

```{r}
METABRIC_RNA_Mutation <- read.csv("METABRIC_RNA_Mutation.csv")
```

```{r}
dat0 <- METABRIC_RNA_Mutation %>%
  as_tibble()

# rename
dat1 <- dat0 %>%
  transmute(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event = overall_survival,
    
    age = age_at_diagnosis,
    tumor_size,
    nodes_positive = lymph_nodes_examined_positive,
    grade = neoplasm_histologic_grade,
    her2 = her2_status,
    pam50 = `pam50_._claudin.low_subtype`,
    tumor_stage,
    menopausal_state = inferred_menopausal_state,
    pr_status,
    surgery = type_of_breast_surgery,
    radio_therapy
  )

# cleaning
dat2 <- dat1 %>%
  mutate(
    across(where(is.character), ~na_if(.x, "")),
    across(c(er_status, hormone_therapy, chemotherapy,
             grade, her2, pam50, tumor_stage,
             menopausal_state, pr_status, surgery, radio_therapy), as.character),
    across(c(age, tumor_size, nodes_positive, os_months, os_event), as.numeric)
  )
```

```{r}
table(dat2$er_status, useNA = "ifany")
table(dat2$hormone_therapy, useNA = "ifany")
table(dat2$chemotherapy, useNA = "ifany")
table(dat2$grade, useNA = "ifany")
table(dat2$her2, useNA = "ifany")
table(dat2$pam50, useNA = "ifany")
table(dat2$tumor_stage, useNA = "ifany")
table(dat2$menopausal_state, useNA = "ifany")
table(dat2$pr_status, useNA = "ifany")
table(dat2$surgery, useNA = "ifany")
table(dat2$radio_therapy, useNA = "ifany")
```
```{r}
dat_table1 <- dat2 %>%
  filter(
    er_status == "Positive",
    hormone_therapy == 1
  ) %>%
  mutate(
    chemo = chemotherapy,
    trt_group = factor(
      chemo,
      levels = c(0, 1),
      labels = c("Hormone only", "Hormone + Chemo")
    )
  ) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event)
  )

nrow(dat0)
nrow(dat_table1)

table(dat_table1$trt_group, useNA = "ifany")
prop.table(table(dat_table1$trt_group))
```
```{r}
vars_check <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status",
  "surgery", "radio_therapy"
)

missing_by_group_long <- dat_table1 %>%
  group_by(trt_group) %>%
  summarise(
    across(all_of(vars_check), ~mean(is.na(.))*100)
  ) %>%
  pivot_longer(
    -trt_group,
    names_to = "variable",
    values_to = "pct_missing"
  )

missing_by_group_long
```

## Table 1: covariate distributions over Trt Groups

```{r}
# 废案
vars <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

factorVars <- c(
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "trt_group",
  data = dat_table1,
  factorVars = factorVars,
  includeNA = TRUE
)

print(
  tab1,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  nonnormal = c("tumor_size", "nodes_positive")
)
```

##### numeric table
Patients receiving chemotherapy were significantly younger (mean age 53.6 vs 66.1 years, p < 0.001) and had more aggressive tumor features, including larger tumor size, greater lymph node involvement, and higher histologic grade (all p < 0.001). Pre-menopausal status was also more common among patients receiving chemotherapy (38.2% vs 8.0%, p < 0.001). HER2 status and PR status did not significantly differ between treatment groups.

These findings suggest that treatment assignment was strongly associated with baseline risk factors.

```{r}
summary_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    g1 <- as.numeric(x[g == "Hormone only"])
    g2 <- as.numeric(x[g == "Hormone + Chemo"])
    
    ttest <- t.test(g1, g2)
    
    mean1 <- mean(g1, na.rm = TRUE)
    sd1 <- sd(g1, na.rm = TRUE)
    mean2 <- mean(g2, na.rm = TRUE)
    sd2 <- sd(g2, na.rm = TRUE)
    smd <- (mean2 - mean1) / sqrt((sd1^2 + sd2^2) / 2)
    
    data.frame(
      Variable = v,
      Mean_Group1 = mean1,
      SD_Group1 = sd1,
      Mean_Group2 = mean2,
      SD_Group2 = sd2,
      t_value = as.numeric(ttest$statistic),
      p_value = ttest$p.value,
      SMD = smd
    )
  })
  
  dplyr::bind_rows(out)
}

cont_vars <- c("age", "tumor_size", "nodes_positive")

cont_table <- summary_table(dat_table1, cont_vars) |>
  dplyr::mutate(
    across(where(is.numeric), ~round(.x, 3))
  )

print(cont_table)
```


##### categorical table
```{r}
categorical_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    # 变成字符，保留 NA 为单独一类
    x_chr <- as.character(x)
    x_chr[is.na(x_chr)] <- "Missing"
    
    g_chr <- as.character(g)
    
    tab <- table(x_chr, g_chr)
    prop <- prop.table(tab, margin = 2) * 100
    
    # 自动选择检验
    test <- if (any(tab < 5)) fisher.test(tab) else chisq.test(tab)
    
    stat_value <- if ("statistic" %in% names(test)) as.numeric(test$statistic) else NA_real_
    p_value <- test$p.value
    
    # 防止列名顺序或缺失导致报错
    n1 <- if ("Hormone only" %in% colnames(tab)) tab[, "Hormone only"] else rep(0, nrow(tab))
    n2 <- if ("Hormone + Chemo" %in% colnames(tab)) tab[, "Hormone + Chemo"] else rep(0, nrow(tab))
    
    p1 <- if ("Hormone only" %in% colnames(prop)) prop[, "Hormone only"] else rep(0, nrow(prop))
    p2 <- if ("Hormone + Chemo" %in% colnames(prop)) prop[, "Hormone + Chemo"] else rep(0, nrow(prop))
    
    df <- data.frame(
      Variable = v,
      Level = rownames(tab),
      n_Group1 = as.numeric(n1),
      pct_Group1 = as.numeric(p1),
      n_Group2 = as.numeric(n2),
      pct_Group2 = as.numeric(p2),
      statistic = NA_real_,
      p_value = NA_real_,
      stringsAsFactors = FALSE
    )
    
    # 只在每个变量第一行放检验结果
    df$statistic[1] <- stat_value
    df$p_value[1] <- p_value
    
    df
  })
  
  dplyr::bind_rows(out)
}

cat_vars <- c(
  "grade",
  "her2",
  "pam50",
  "tumor_stage",
  "menopausal_state",
  "pr_status",
  "surgery",
  "radio_therapy"
)

cat_table <- categorical_table(dat_table1, cat_vars) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~round(.x, 3))
  )

print(cat_table)
```




## gene variable selection: LASSO/SCAD/MCP/coxBAR
Apply the following penalized regression methods for variable selection:

• LASSO (Least Absolute Shrinkage and Selection Operator)
• BAR (Broken Adaptive Ridge)
• SCAD (Smoothly Clipped Absolute Deviation)
• MCP (Minimax Concave Penalty)
```{r}
library(ncvreg)
```
### set up dataset for penalized regression
331 gene expression z-scores: numeric
175 gene mutation variables: either 0 or "strings", which represents whether the gene mutates and the mutation type. Strings cannot pass into cox penalized regression due to categorical high-cardinality, and each mutation type have very few samples, so penalized cox regression cannot be able to handle them. To address this problem, we would re-encode them to:
= 1 if mutated, or = 0 if no mutation.


We would also restrict the sample to ER+ and hormone_therapy ==1 only, same as our analysis cohort.
```{r}
dat_gene0 <- dat0 %>%
  select(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event = overall_survival,
    brca1:siah1_mut
  )

dat_gene1 <- dat_gene0 %>%
  mutate(
    across(where(is.character), ~na_if(.x, "")),
    across(c(er_status, hormone_therapy, chemotherapy), as.character),
    across(c(os_months, os_event), as.numeric),
    across(brca1:ugt2b7, as.numeric),
    across(
      pik3ca_mut:siah1_mut,
      ~case_when(
        is.na(.x) ~ NA_real_,
        .x == "0" ~ 0,
        TRUE ~ 1
      )
    )
  )

dat_lasso0 <- dat_gene1 %>%
  filter(
    er_status == "Positive",
    hormone_therapy == 1
  ) %>%
  filter(
    !is.na(chemotherapy),
    !is.na(os_months),
    !is.na(os_event)
  )
```

```{r}
gene_df <- dat_lasso0 %>%
  select(brca1:siah1_mut)

col_na <- colSums(is.na(gene_df))

col_var0 <- sapply(gene_df, function(x) var(x, na.rm = TRUE))
table(col_var0 == 0)

keep_gene <- names(gene_df)[col_na == 0 & col_var0 > 0]

dat_lasso <- dat_lasso0 %>%
  select(patient_id, er_status, hormone_therapy, chemotherapy, os_months, os_event,
         all_of(keep_gene))
```

```{r}
X <- dat_lasso %>%
  select(all_of(keep_gene)) %>%
  as.data.frame() %>%
  as.matrix()

sum(is.na(X)) # should be 0
dim(X) # should be 662-2 = 660
X <- scale(X)
```
### LASSO
```{r}
library(ncvreg)

set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "lasso",
  nlambda = 50,
  nfolds = 10
)
```
```{r}
fit.lasso <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "lasso",
  lambda = fit$lambda.min
)

fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```
selected:
```{r}
lasso_coef <- coef(fit.lasso)

lasso_selected <- names(lasso_coef)[lasso_coef != 0]

lasso_selected

lasso_selected_df <- data.frame(
  variable = names(lasso_coef)[lasso_coef != 0],
  coef = lasso_coef[lasso_coef != 0]
) %>%
  arrange(desc(abs(coef)))

lasso_selected_df
```

### SCAD
```{r}
set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "SCAD",
  nlambda = 50,
  nfolds = 10
)

# check ups (avoid the case if only a few fits worked)
fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```

```{r, warning=FALSE}
fit.scad <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "SCAD",
  lambda = fit$lambda.min
)
```

```{r}
scad_coef <- coef(fit.scad)

scad_selected <- names(scad_coef)[scad_coef != 0]

scad_selected

scad_selected_df <- data.frame(
  variable = names(scad_coef)[scad_coef != 0],
  coef = scad_coef[scad_coef != 0]
) %>%
  arrange(desc(abs(coef)))

scad_selected_df
```

### MCP
```{r}
set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "MCP",
  nlambda = 50,
  nfolds = 10
)

fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```

```{r, warning=FALSE}
fit.mcp <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "MCP",
  lambda = fit$lambda.min
)
```

```{r}
mcp_coef <- coef(fit.mcp)

mcp_selected <- names(mcp_coef)[mcp_coef != 0]

mcp_selected

mcp_selected_df <- data.frame(
  variable = names(mcp_coef)[mcp_coef != 0],
  coef = mcp_coef[mcp_coef != 0]
) %>%
  arrange(desc(abs(coef)))

mcp_selected_df
```

### coxBAR
```{r}
library(Cyclops)
library(BrokenAdaptiveRidge)
```

```{r}
library(Cyclops)
library(BrokenAdaptiveRidge)
library(survival)

set.seed(215)

ncov <- ncol(X)
n <- nrow(X)

dataFit <- createCyclopsData(
  Surv(dat_lasso$os_months, dat_lasso$os_event) ~ X,
  modelType = "cox"
)

pen_grid <- exp(seq(log(0.1 * log(ncov)), log(1 * log(ncov)), length.out = 50))

coxbar_grid <- lapply(pen_grid, function(pen) {

  barPrior <- createBarPrior(
    penalty = pen,
    initialRidgeVariance = 1 / pen
  )

  fit.bar <- try(
    fitCyclopsModel(dataFit, prior = barPrior),
    silent = TRUE
  )

  if(inherits(fit.bar, "try-error")) {
    return(list(
      penalty = pen,
      fit = NULL,
      bic = Inf,
      refit_warn = FALSE,
      refit_ok = FALSE
    ))
  }

  coef_bar <- try(coef(fit.bar), silent = TRUE)

  if(inherits(coef_bar, "try-error") || is.null(coef_bar)) {
    return(list(
      penalty = pen,
      fit = NULL,
      bic = Inf,
      refit_warn = FALSE,
      refit_ok = FALSE
    ))
  }

  selected_idx <- which(coef_bar != 0)
  k <- length(selected_idx)

  if(k == 0) {
    fit_null <- coxph(Surv(os_months, os_event) ~ 1, data = dat_lasso)

    bic <- -2 * as.numeric(logLik(fit_null)) + log(n) * k

    return(list(
      penalty = pen,
      fit = fit.bar,
      bic = bic,
      refit_warn = FALSE,
      refit_ok = TRUE
    ))
  } else {

    dat_bar <- data.frame(
      os_months = dat_lasso$os_months,
      os_event = dat_lasso$os_event,
      X[, selected_idx, drop = FALSE]
    )

    colnames(dat_bar)[3:ncol(dat_bar)] <- colnames(X)[selected_idx]

    warn_msg <- NULL

    fit_refit <- tryCatch(
      withCallingHandlers(
        coxph(
          as.formula(
            paste(
              "Surv(os_months, os_event) ~",
              paste(colnames(X)[selected_idx], collapse = " + ")
            )
          ),
          data = dat_bar
        ),
        warning = function(w) {
          warn_msg <<- conditionMessage(w)
          invokeRestart("muffleWarning")
        }
      ),
      error = function(e) NULL
    )

    if(is.null(fit_refit)) {
      bic <- Inf
      refit_ok <- FALSE
    } else {
      bic <- -2 * as.numeric(logLik(fit_refit)) + log(n) * k
      refit_ok <- TRUE
    }

    return(list(
      penalty = pen,
      fit = fit.bar,
      bic = bic,
      refit_warn = !is.null(warn_msg),
      warn_msg = warn_msg,
      refit_ok = refit_ok
    ))
  }
})
```

```{r}
bic_values <- sapply(coxbar_grid, function(x) x$bic)

best_index <- which.min(bic_values)

pen_grid[best_index]
range(pen_grid)

best_index
length(pen_grid)
```

best fit:
```{r}
fit.coxbar <- coxbar_grid[[best_index]]$fit

coef_bar <- coef(fit.coxbar)
```

selected vars:
```{r}
coxbar_selected <- colnames(X)[coef_bar != 0]

coxbar_selected

coxbar_selected_df <- data.frame(
  variable = colnames(X)[coef_bar != 0],
  coef = coef_bar[coef_bar != 0]
) %>%
  arrange(desc(abs(coef)))

coxbar_selected_df
```

```{r}
sum(is.infinite(bic_values))

sum(sapply(coxbar_grid, function(x) x$refit_warn), na.rm = TRUE)

sum(sapply(coxbar_grid, function(x) x$refit_ok), na.rm = TRUE)
```
only 2 runs failed out of 50 penalties, all good;
only 1 run had infinite coefficient warning, ignorable;
48 runs of different penalties have good refit Cox models.

### compare methods
```{r}
gene_sets <- list(
  lasso = lasso_selected,
  scad = scad_selected,
  mcp = mcp_selected,
  coxbar = coxbar_selected
)
```

```{r}
refit_compare <- lapply(names(gene_sets), function(m) {

  vars <- gene_sets[[m]]

  dat_tmp <- dat_lasso %>%
    select(os_months, os_event, all_of(vars))

  fit_tmp <- coxph(
    as.formula(
      paste("Surv(os_months, os_event) ~", paste(vars, collapse = " + "))
    ),
    data = dat_tmp
  )

  data.frame(
    method = m,
    n_var = length(vars),
    logLik = as.numeric(logLik(fit_tmp)),
    AIC = AIC(fit_tmp),
    BIC = BIC(fit_tmp)
  )
}) %>%
  bind_rows() %>%
  arrange(BIC)

refit_compare
```


### select vars
Our aim is to select 5~8 final gene variables to enter the cox model.


why choose coxBAR as main model (除了BIC最小最好之外):
Because our goal was sparse and stable variable screening in a correlated genomic setting, we treated CoxBAR as the primary penalized Cox selection method. CoxBAR is a surrogate ℓ0 penalized Cox approach based on broken adaptive ridge and has been shown to possess selection consistency, oracle property for parameter estimation, and a grouping property for highly correlated covariates. In simulation and empirical evaluations, BAR based Cox regression selected relatively sparse models and achieved model selection and discriminatory performance comparable to or slightly better than Cox LASSO, with less biased and more stable solution paths under BIC based tuning.

Summarize selected variables overlapping:
```{r}
all_selected <- list(
  lasso = lasso_selected,
  scad = scad_selected,
  mcp = mcp_selected,
  coxbar = coxbar_selected
)

all_genes <- sort(unique(unlist(all_selected)))

overlap_df <- data.frame(variable = all_genes) %>%
  mutate(
    in_lasso = variable %in% lasso_selected,
    in_scad = variable %in% scad_selected,
    in_mcp = variable %in% mcp_selected,
    in_coxbar = variable %in% coxbar_selected,
    n_methods = in_lasso + in_scad + in_mcp + in_coxbar
  ) %>%
  arrange(desc(n_methods), variable)

overlap_df
```


在 4 个方法都出现 的变量是：
tsc2, bbc3, sbno1, ptprd_mut, nek1_mut

再从 coxBAR 中那些 出现了 3 次 的变量里补 1 到 2 个，例如：

mapk14, kmt2c, casp8, hsd17b4

这样最后凑成 6 到 7 个：
tsc2, bbc3, sbno1, ptprd_mut, nek1_mut, mapk14, kmt2c (coxBAR里最重要的那两个)

这组有两个优点。第一，完全以 coxBAR 为主。第二，稳定性很强，因为这 7 个里前 5 个在所有方法都出现，后 2 个也在 3 个方法里出现。


最后选出来的变量：tsc2, bbc3, sbno1, ptprd_mut, nek1_mut, mapk14, kmt2c。注意ptprd_mut和 nek1_mut变量的原始格式是0或string，要把string全变成1，代表有突变，0代表无突变，再加入我们的cox model作为协变量。



