---
title: "Project part Total"
author: "Hanxi Chen, Hanying Xu, Xiaomiao Wang"
subtitle: :)
date: today
format:
  pdf:
    documentclass: article
    number-sections: false
    fig-align: center
    keep-tex: false  # keeps the intermediate .tex file
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    html-math-method: mathjax
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r, warning=FALSE}
library(tidyverse)
library(car)
library(dplyr)
library(survival)
library(survminer)
library(zoo) # na.locf

library(janitor)
library(skimr)

library(tableone)

library(tableone)
library(gtsummary)
```


## Import and Cleaning

```{r}
METABRIC_RNA_Mutation <- read.csv("METABRIC_RNA_Mutation.csv")
```


```{r}
dat0 <- METABRIC_RNA_Mutation %>%
  as_tibble()

# rename variables
dat1 <- dat0 %>%
  transmute(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event = overall_survival,
    
    age = age_at_diagnosis,
    tumor_size,
    nodes_positive = lymph_nodes_examined_positive,
    grade = neoplasm_histologic_grade,
    her2 = her2_status,
    pam50 = `pam50_._claudin.low_subtype`,
    tumor_stage,
    menopausal_state = inferred_menopausal_state,
    pr_status,
    surgery = type_of_breast_surgery,
    radio_therapy
  )

# cleaning variables
dat2 <- dat1 %>%
  mutate(
    across(where(is.character), ~na_if(.x, "")),
    across(c(er_status, hormone_therapy, chemotherapy,
             grade, her2, pam50, tumor_stage,
             menopausal_state, pr_status, surgery, radio_therapy), as.character),
    across(c(age, tumor_size, nodes_positive, os_months, os_event), as.numeric)
  )
```

```{r}
table(dat2$er_status, useNA = "ifany")
table(dat2$hormone_therapy, useNA = "ifany")
table(dat2$chemotherapy, useNA = "ifany")
table(dat2$grade, useNA = "ifany")
table(dat2$her2, useNA = "ifany")
table(dat2$pam50, useNA = "ifany")
table(dat2$tumor_stage, useNA = "ifany")
table(dat2$menopausal_state, useNA = "ifany")
table(dat2$pr_status, useNA = "ifany")
table(dat2$surgery, useNA = "ifany")
table(dat2$radio_therapy, useNA = "ifany")
```
```{r}
dat_table1 <- dat2 %>%
  filter(
    er_status == "Positive",
    hormone_therapy == 1
  ) %>%
  mutate(
    chemo = chemotherapy,
    trt_group = factor(
      chemo,
      levels = c(0, 1),
      labels = c("Hormone only", "Hormone + Chemo")
    )
  ) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event)
  )

nrow(dat0)
nrow(dat_table1)

table(dat_table1$trt_group, useNA = "ifany")
prop.table(table(dat_table1$trt_group))
```
```{r}
vars_check <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status",
  "surgery", "radio_therapy"
)

missing_by_group_long <- dat_table1 %>%
  group_by(trt_group) %>%
  summarise(
    across(all_of(vars_check), ~mean(is.na(.))*100)
  ) %>%
  pivot_longer(
    -trt_group,
    names_to = "variable",
    values_to = "pct_missing"
  )

missing_by_group_long
```



## Table 1: covariate distributions over Trt Groups

```{r}
# raw version
vars <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

factorVars <- c(
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "trt_group",
  data = dat_table1,
  factorVars = factorVars,
  includeNA = TRUE
)

print(
  tab1,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  nonnormal = c("tumor_size", "nodes_positive")
)
```

##### numeric table
Patients receiving chemotherapy were significantly younger (mean age 53.6 vs 66.1 years, p < 0.001) and had more aggressive tumor features, including larger tumor size, greater lymph node involvement, and higher histologic grade (all p < 0.001). Pre-menopausal status was also more common among patients receiving chemotherapy (38.2% vs 8.0%, p < 0.001). HER2 status and PR status did not significantly differ between treatment groups.

These findings suggest that treatment assignment was strongly associated with baseline risk factors.

```{r}
summary_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    g1 <- as.numeric(x[g == "Hormone only"])
    g2 <- as.numeric(x[g == "Hormone + Chemo"])
    
    ttest <- t.test(g1, g2)
    
    mean1 <- mean(g1, na.rm = TRUE)
    sd1 <- sd(g1, na.rm = TRUE)
    mean2 <- mean(g2, na.rm = TRUE)
    sd2 <- sd(g2, na.rm = TRUE)
    smd <- (mean2 - mean1) / sqrt((sd1^2 + sd2^2) / 2)
    
    data.frame(
      Variable = v,
      Mean_Group1 = mean1,
      SD_Group1 = sd1,
      Mean_Group2 = mean2,
      SD_Group2 = sd2,
      t_value = as.numeric(ttest$statistic),
      p_value = ttest$p.value,
      SMD = smd
    )
  })
  
  dplyr::bind_rows(out)
}

cont_vars <- c("age", "tumor_size", "nodes_positive")

cont_table <- summary_table(dat_table1, cont_vars) |>
  dplyr::mutate(
    across(where(is.numeric), ~round(.x, 3))
  )

print(cont_table)
```


##### categorical table
```{r}
categorical_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    # convert to character and keep missing as a level
    x_chr <- as.character(x)
    x_chr[is.na(x_chr)] <- "Missing"
    
    g_chr <- as.character(g)
    
    tab <- table(x_chr, g_chr)
    prop <- prop.table(tab, margin = 2) * 100
    
    # choose test automatically
    test <- if (any(tab < 5)) fisher.test(tab) else chisq.test(tab)
    
    stat_value <- if ("statistic" %in% names(test)) as.numeric(test$statistic) else NA_real_
    p_value <- test$p.value
    
    # avoid errors from missing group columns
    n1 <- if ("Hormone only" %in% colnames(tab)) tab[, "Hormone only"] else rep(0, nrow(tab))
    n2 <- if ("Hormone + Chemo" %in% colnames(tab)) tab[, "Hormone + Chemo"] else rep(0, nrow(tab))
    
    p1 <- if ("Hormone only" %in% colnames(prop)) prop[, "Hormone only"] else rep(0, nrow(prop))
    p2 <- if ("Hormone + Chemo" %in% colnames(prop)) prop[, "Hormone + Chemo"] else rep(0, nrow(prop))
    
    df <- data.frame(
      Variable = v,
      Level = rownames(tab),
      n_Group1 = as.numeric(n1),
      pct_Group1 = as.numeric(p1),
      n_Group2 = as.numeric(n2),
      pct_Group2 = as.numeric(p2),
      statistic = NA_real_,
      p_value = NA_real_,
      stringsAsFactors = FALSE
    )
    
    # put test results on the first row only
    df$statistic[1] <- stat_value
    df$p_value[1] <- p_value
    
    df
  })
  
  dplyr::bind_rows(out)
}

cat_vars <- c(
  "grade",
  "her2",
  "pam50",
  "tumor_stage",
  "menopausal_state",
  "pr_status",
  "surgery",
  "radio_therapy"
)

cat_table <- categorical_table(dat_table1, cat_vars) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~round(.x, 3))
  )

print(cat_table)
```

### RMST
```{r}
ps_vars <- c(
  "chemo",
  "trt_group",
  "os_months",
  "os_event",
  "age",
  "tumor_size",
  "nodes_positive",
  "grade",
  "pam50"
)

dat_km <- dat_table1 %>%
  select(all_of(ps_vars)) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50)
  ) %>%
  mutate(
    chemo = as.numeric(chemo)
  )
```

```{r}
library(survRM2)
rmst_res <- rmst2(
  time   = dat_km$os_months,
  status = dat_km$os_event,
  arm    = dat_km$chemo,
  tau    = 100
)
rmst_res
```



## Crude KM & Log-rank test: unadjusted association
### KM curve
The Kaplan–Meier estimator was used to estimate the survival function for each treatment group.
The log-rank test was used to compare the survival distributions between treatment groups.
```{r}
library(survival)
library(survminer)
```

```{r}
ps_vars <- c(
  "chemo",
  "trt_group",
  "os_months",
  "os_event",
  "age",
  "tumor_size",
  "nodes_positive",
  "grade",
  "pam50"
)

dat_km <- dat_table1 %>%
  select(all_of(ps_vars)) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50)
  ) %>%
  mutate(
    chemo = as.numeric(chemo)
  )
  


table(dat_km$trt_group)
prop.table(table(dat_km$trt_group))
```

```{r}
tau <- 100

fit_km <- survfit(
  Surv(os_months, os_event) ~ trt_group,
  data = dat_km
)

ggsurvplot(
  fit_km,
  data = dat_km,
  risk.table = TRUE,
  pval = TRUE,
  conf.int = TRUE,
  xlab = "Time since diagnosis (months)",
  ylab = "Overall survival probability",
  legend.title = "Treatment group",
  legend.labs = c("Hormone only", "Hormone + Chemo"),
  xlim = c(0, tau),
  break.time.by = 25,
  ggtheme = theme_bw()
)

summary(fit_km, times = tau)
```

The crude Kaplan-Meier curves suggested worse survival in the chemotherapy group. However, this comparison is unadjusted and may reflect baseline differences in disease severity rather than treatment effect.

```{r}
survdiff(Surv(os_months, os_event) ~ trt_group, data = dat_km)
```


## Weighted KM and Log-rank Test
### calculate propensity score
set up dataset for Propensity score models:
```{r}
ps_vars <- c(
  "chemo",
  "trt_group",
  "os_months",
  "os_event",
  "age",
  "tumor_size",
  "nodes_positive",
  "grade",
  "pam50"
)

dat_ps <- dat_table1 %>%
  select(all_of(ps_vars)) %>%
  mutate(
    pam50_grp = case_when(
      pam50 %in% c("LumA", "LumB") ~ pam50,
      TRUE ~ "Other"
    )
  ) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50_grp)
  ) %>%
  mutate(
    chemo = as.numeric(chemo),
    grade = factor(grade),
    pam50_grp = factor(pam50_grp, levels = c("LumA", "LumB", "Other"))
  )
  

nrow(dat_table1)
nrow(dat_ps)

table(dat_ps$trt_group)
prop.table(table(dat_ps$trt_group))
```

### fit propensity score model and calculate IPTW
A logistic regression model was used to estimate the propensity score for receiving chemotherapy.
```{r}
ps_fit <- glm(
  chemo ~ age + tumor_size + nodes_positive + grade + pam50_grp,
  data = dat_ps,
  family = binomial()
)

summary(ps_fit)
```

calculate stabilized IPTW:
```{r}
dat_ps <- dat_ps %>%
  mutate(
    ps = predict(ps_fit, type = "response"),
    p_treat = mean(chemo == 1),
    sw = if_else(
      chemo == 1,
      p_treat / ps,
      (1 - p_treat) / (1 - ps)
    ),
    iptw = if_else(
      chemo == 1,
      1 / ps,
      1 / (1 - ps)
    )
  )
```

### check overlap
```{r}
dat_ps %>%
  group_by(trt_group) %>%
  summarise(
    n = n(),
    min_ps = min(ps),
    q1_ps = quantile(ps, 0.25),
    median_ps = median(ps),
    mean_ps = mean(ps),
    q3_ps = quantile(ps, 0.75),
    max_ps = max(ps)
  )
```

```{r}
ggplot(dat_ps, aes(x = ps, fill = trt_group)) +
  geom_density(alpha = 0.35) +
  labs(
    x = "Propensity score",
    y = "Density",
    fill = "Treatment group"
  ) +
  theme_bw()
```


```{r}
ps_range <- dat_ps %>%
  group_by(trt_group) %>%
  summarise(
    min_ps = min(ps),
    max_ps = max(ps)
  )

ps_range
```

check weight distribution:
```{r}
summary(dat_ps$sw)
quantile(dat_ps$sw, probs = c(0, 0.01, 0.05, 0.25, 0.5, 0.75, 0.95, 0.99, 1))
```

```{r}
ggplot(dat_ps, aes(x = sw, fill = trt_group)) +
  geom_histogram(position = "identity", alpha = 0.35, bins = 30) +
  labs(
    x = "Stabilized IPTW",
    y = "Count",
    fill = "Treatment group"
  ) +
  theme_bw()
```

balance before weighting:
```{r}
vars_balance <- c(
  "age",
  "tumor_size",
  "nodes_positive",
  "grade",
  "pam50_grp"
)

factorVars_balance <- c(
  "grade",
  "pam50_grp"
)

tab_unw <- CreateTableOne(
  vars = vars_balance,
  strata = "trt_group",
  data = dat_ps,
  factorVars = factorVars_balance,
  includeNA = TRUE,
  smd = TRUE
)

print(
  tab_unw,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  smd = TRUE,
  nonnormal = c("tumor_size", "nodes_positive")
)
```

balance after weighting:
```{r}
library(survey)

design_sw <- svydesign(
  ids = ~1,
  weights = ~sw,
  data = dat_ps
)

tab_w <- svyCreateTableOne(
  vars = vars_balance,
  strata = "trt_group",
  data = design_sw,
  factorVars = factorVars_balance,
  includeNA = TRUE,
  smd = TRUE
)

print(
  tab_w,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  smd = TRUE,
  nonnormal = c("tumor_size", "nodes_positive")
)
```
### compare SMD before and after weighting
SMD < 0.1 → ideal
SMD < 0.2 → acceptable
SMD > 0.2 → imbalance
```{r}
smd_unw <- ExtractSmd(tab_unw)
smd_w <- ExtractSmd(tab_w)

smd_compare <- data.frame(
  variable = vars_balance,
  smd_unweighted = as.numeric(smd_unw),
  smd_weighted = as.numeric(smd_w)
) %>%
  arrange(desc(smd_unweighted))

smd_compare
```

```{r}
smd_compare %>%
  mutate(
    imbalance_unweighted = abs(smd_unweighted) > 0.1,
    imbalance_weighted = abs(smd_weighted) > 0.1
  )
```

We estimated a propensity score for receiving chemotherapy using baseline clinical covariates age, tumor size, number of positive lymph nodes, histologic grade, and PAM50 subtype, then constructed stabilized inverse probability of treatment weights. The propensity score distributions showed adequate overlap between treatment groups, although the chemotherapy group had generally larger scores, consistent with confounding by indication. The stabilized weights were overall well behaved, with mean close to 1 and only a few large values, suggesting that the weighted analysis should be reasonably stable. Covariate balance improved meaningfully after weighting, especially for nodal involvement and tumor size, while some residual imbalance remained for age, grade, and PAM50. Overall, these diagnostics suggest that IPTW is appropriate as an adjusted survival comparison, while residual confounding should still be acknowledged.


### IPTW weighted KM
```{r}
# fit weighted km curves
fit_km_w <- survfit(
  Surv(os_months, os_event) ~ trt_group,
  data = dat_ps,
  weights = sw
)

ggsurvplot(
  fit_km_w,
  data = dat_ps,
  conf.int = TRUE,
  xlab = "Time since diagnosis (months)",
  ylab = "Overall survival probability",
  legend.title = "Treatment group",
  legend.labs = c("Hormone only", "Hormone + Chemo"),
  xlim = c(0, tau),
  break.time.by = 25,
  ggtheme = theme_bw()
)

summary(fit_km_w, times = tau)

```


```{r}
kmw_150 <- summary(fit_km_w, times = tau)

data.frame(
  trt_group = kmw_150$strata,
  time = kmw_150$time,
  surv = kmw_150$surv,
  lower = kmw_150$lower,
  upper = kmw_150$upper
)
```


```{r}
ggsurvplot(
  fit_km_w,
  data = dat_ps,
  conf.int = TRUE,
  title = "Weighted Kaplan–Meier Survival Curves by Treatment Group",
  xlab = "Time since diagnosis (months)",
  ylab = "Overall survival probability",
  legend.title = "Treatment group",
  legend.labs = c("Hormone only", "Hormone + Chemo"),
  xlim = c(0, tau),
  break.time.by = 25,
  ggtheme = theme_bw()
)
```

### IPTW weighted logrank
```{r}
design_sw <- svydesign(
  ids = ~1,
  weights = ~sw,
  data = dat_ps
)
```

```{r}
library(survey)

fit_wlr <- svylogrank(
  Surv(os_months, os_event) ~ trt_group,
  design = design_sw
)

fit_wlr
```

### IPTW weighted cox
```{r}
dat_wcox <- dat_table1 %>%
  select(
    patient_id,
    chemo,
    trt_group,
    os_months,
    os_event,
    age,
    tumor_size,
    nodes_positive,
    grade,
    pam50
  ) %>%
  mutate(
    pam50_grp = case_when(
      pam50 %in% c("LumA", "LumB") ~ pam50,
      TRUE ~ "Other"
    )
  ) %>%
  filter(
    !is.na(patient_id),
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50_grp)
  ) %>%
  mutate(
    chemo = as.numeric(chemo),
    grade = factor(grade),
    pam50_grp = factor(pam50_grp, levels = c("LumA", "LumB", "Other"))
  )
```


```{r}
ps_fit <- glm(
  chemo ~ age + tumor_size + nodes_positive + grade + pam50_grp,
  data = dat_wcox,
  family = binomial()
)

dat_wcox <- dat_wcox %>%
  mutate(
    ps = predict(ps_fit, type = "response"),
    p_treat = mean(chemo == 1),
    sw = if_else(
      chemo == 1,
      p_treat / ps,
      (1 - p_treat) / (1 - ps)
    )
  )
```

#### weighted cox without time-dependent chemotherapy:
```{r}
cox_w <- coxph(
  Surv(os_months, os_event) ~
    chemo +
    age +
    tumor_size +
    nodes_positive +
    grade +
    pam50_grp,
  data = dat_wcox,
  weights = sw,
  ties = "breslow",
  robust = TRUE,
  id = patient_id
)

summary(cox_w)
```

```{r}
ph_test_w <- cox.zph(cox_w)
ph_test_w
```

```{r}
event.times <- sort(unique(dat_wcox$os_months[dat_wcox$os_event == 1]))
event.times <- event.times[-1]

dat_split_w <- survSplit(
  data = dat_wcox,
  cut = event.times,
  end = "os_months",
  start = "t0",
  event = "os_event"
)
```

```{r}
dat_split_w <- dat_split_w %>%
  mutate(
    chemo_t = chemo * log(os_months)
  )
```

#### weighted cox with time-dependent chemotherapy
```{r}
cox_w_td <- coxph(
  Surv(t0, os_months, os_event) ~
    chemo +
    chemo_t +
    age +
    tumor_size +
    nodes_positive +
    grade +
    pam50_grp,
  data = dat_split_w,
  weights = sw,
  ties = "breslow",
  robust = TRUE,
  id = patient_id
)

summary(cox_w_td)
```
#### clinical-only crude Cox
```{r}
# set up data for clinical-only crude cox
dat_ccox <- dat_table1 %>%
  select(
    patient_id,
    chemo,
    trt_group,
    os_months,
    os_event,
    age,
    tumor_size,
    nodes_positive,
    grade,
    pam50
  ) %>%
  mutate(
    pam50_grp = case_when(
      pam50 %in% c("LumA", "LumB") ~ pam50,
      TRUE ~ "Other"
    )
  ) %>%
  filter(
    !is.na(patient_id),
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50_grp)
  ) %>%
  mutate(
    chemo = factor(
      as.numeric(chemo),
      levels = c(0, 1),
      labels = c("Hormone only", "Hormone + Chemo")
    ),
    chemo_num = as.numeric(chemo == "Hormone + Chemo"),
    grade = factor(grade),
    pam50_grp = factor(pam50_grp, levels = c("LumA", "LumB", "Other"))
  )

nrow(dat_ccox)
table(dat_ccox$chemo, useNA = "ifany")
table(dat_ccox$grade, useNA = "ifany")
table(dat_ccox$pam50_grp, useNA = "ifany")
```


```{r}
# fit clinical-only crude cox
ccox_fit <- coxph(
  Surv(os_months, os_event) ~
    chemo +
    age +
    tumor_size +
    nodes_positive +
    grade +
    pam50_grp,
  data = dat_ccox,
  ties = "breslow"
)

summary(ccox_fit)
```

```{r}
# check ph assumption
ccox_ph_res <- cox.zph(ccox_fit)
ccox_ph_res
plot(ccox_ph_res)
```

#### clinical-only time-dependent cox
```{r}
# set up time-dependent crude cox
ccox_event_times <- sort(unique(dat_ccox$os_months[dat_ccox$os_event == 1]))
ccox_event_times <- ccox_event_times[-1]

dat_ccox_split <- survSplit(
  data = dat_ccox,
  cut = ccox_event_times,
  end = "os_months",
  start = "t0",
  event = "os_event"
) %>%
  mutate(
    chemo_t = chemo_num * log(os_months)
  )
```

```{r}
# fit time-dependent crude cox
ccox_td_fit <- coxph(
  Surv(t0, os_months, os_event) ~
    chemo_num +
    chemo_t +
    age +
    tumor_size +
    nodes_positive +
    grade +
    pam50_grp,
  data = dat_ccox_split,
  ties = "breslow"
)

summary(ccox_td_fit)
```

```{r}
# check ph for td crude cox
ccox_td_ph_res <- cox.zph(ccox_td_fit)
ccox_td_ph_res
plot(ccox_td_ph_res)
```

## Gene variable selection: LASSO/SCAD/MCP/coxBAR
The following penalized regression methods were applied for gene selection:

• LASSO (Least Absolute Shrinkage and Selection Operator)
• BAR (Broken Adaptive Ridge)
• SCAD (Smoothly Clipped Absolute Deviation)
• MCP (Minimax Concave Penalty)
```{r}
library(ncvreg)
```

### set up data for penalized regression
The gene data include 331 gene expression z-scores and 175 mutation variables. The expression variables are numeric, whereas the mutation variables are recorded as either 0 or mutation-type strings. Because mutation-type strings are too sparse and high-dimensional for penalized Cox regression, the mutation variables were recoded as binary indicators, with 1 indicating mutation and 0 indicating no mutation.

The analysis was restricted to ER-positive patients who received hormone therapy, consistent with the main study cohort.


We would also restrict the sample to ER+ and hormone_therapy ==1 only, same as our analysis cohort.
```{r}
dat_gene0 <- dat0 %>%
  select(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event = overall_survival,
    brca1:siah1_mut
  )

dat_gene1 <- dat_gene0 %>%
  mutate(
    across(where(is.character), ~na_if(.x, "")),
    across(c(er_status, hormone_therapy, chemotherapy), as.character),
    across(c(os_months, os_event), as.numeric),
    across(brca1:ugt2b7, as.numeric),
    across(
      pik3ca_mut:siah1_mut,
      ~case_when(
        is.na(.x) ~ NA_real_,
        .x == "0" ~ 0,
        TRUE ~ 1
      )
    )
  )

dat_lasso0 <- dat_gene1 %>%
  filter(
    er_status == "Positive",
    hormone_therapy == 1
  ) %>%
  filter(
    !is.na(chemotherapy),
    !is.na(os_months),
    !is.na(os_event)
  )
```

```{r}
gene_df <- dat_lasso0 %>%
  select(brca1:siah1_mut)

col_na <- colSums(is.na(gene_df))

col_var0 <- sapply(gene_df, function(x) var(x, na.rm = TRUE))
table(col_var0 == 0)

keep_gene <- names(gene_df)[col_na == 0 & col_var0 > 0]

dat_lasso <- dat_lasso0 %>%
  select(patient_id, er_status, hormone_therapy, chemotherapy, os_months, os_event,
         all_of(keep_gene))
```

```{r}
X <- dat_lasso %>%
  select(all_of(keep_gene)) %>%
  as.data.frame() %>%
  as.matrix()

sum(is.na(X)) # should be 0
dim(X) # should be 662-2 = 660
X <- scale(X)
```
### LASSO
```{r}
library(ncvreg)

set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "lasso",
  nlambda = 50,
  nfolds = 10
)
```
```{r}
fit.lasso <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "lasso",
  lambda = fit$lambda.min
)

fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```
selected variables:
```{r}
lasso_coef <- coef(fit.lasso)

lasso_selected <- names(lasso_coef)[lasso_coef != 0]

lasso_selected

lasso_selected_df <- data.frame(
  variable = names(lasso_coef)[lasso_coef != 0],
  coef = lasso_coef[lasso_coef != 0]
) %>%
  arrange(desc(abs(coef)))

lasso_selected_df
```

### SCAD
```{r}
set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "SCAD",
  nlambda = 50,
  nfolds = 10
)

# check ups (avoid the case if only a few fits worked)
fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```

```{r, warning=FALSE}
fit.scad <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "SCAD",
  lambda = fit$lambda.min
)
```

```{r}
scad_coef <- coef(fit.scad)

scad_selected <- names(scad_coef)[scad_coef != 0]

scad_selected

scad_selected_df <- data.frame(
  variable = names(scad_coef)[scad_coef != 0],
  coef = scad_coef[scad_coef != 0]
) %>%
  arrange(desc(abs(coef)))

scad_selected_df
```

### MCP
```{r}
set.seed(215)

fit <- cv.ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "MCP",
  nlambda = 50,
  nfolds = 10
)

fit$lambda.min
range(fit$lambda)

which.min(abs(fit$lambda - fit$lambda.min))
length(fit$lambda)
```

```{r, warning=FALSE}
fit.mcp <- ncvsurv(
  X,
  Surv(dat_lasso$os_months, dat_lasso$os_event),
  penalty = "MCP",
  lambda = fit$lambda.min
)
```

```{r}
mcp_coef <- coef(fit.mcp)

mcp_selected <- names(mcp_coef)[mcp_coef != 0]

mcp_selected

mcp_selected_df <- data.frame(
  variable = names(mcp_coef)[mcp_coef != 0],
  coef = mcp_coef[mcp_coef != 0]
) %>%
  arrange(desc(abs(coef)))

mcp_selected_df
```

### coxBAR
```{r}
library(Cyclops)
library(BrokenAdaptiveRidge)
```

```{r}
library(Cyclops)
library(BrokenAdaptiveRidge)
library(survival)

set.seed(215)

ncov <- ncol(X)
n <- nrow(X)

dataFit <- createCyclopsData(
  Surv(dat_lasso$os_months, dat_lasso$os_event) ~ X,
  modelType = "cox"
)

pen_grid <- exp(seq(log(0.1 * log(ncov)), log(1 * log(ncov)), length.out = 50))

coxbar_grid <- lapply(pen_grid, function(pen) {

  barPrior <- createBarPrior(
    penalty = pen,
    initialRidgeVariance = 1 / pen
  )

  fit.bar <- try(
    fitCyclopsModel(dataFit, prior = barPrior),
    silent = TRUE
  )

  if(inherits(fit.bar, "try-error")) {
    return(list(
      penalty = pen,
      fit = NULL,
      bic = Inf,
      refit_warn = FALSE,
      refit_ok = FALSE
    ))
  }

  coef_bar <- try(coef(fit.bar), silent = TRUE)

  if(inherits(coef_bar, "try-error") || is.null(coef_bar)) {
    return(list(
      penalty = pen,
      fit = NULL,
      bic = Inf,
      refit_warn = FALSE,
      refit_ok = FALSE
    ))
  }

  selected_idx <- which(coef_bar != 0)
  k <- length(selected_idx)

  if(k == 0) {
    fit_null <- coxph(Surv(os_months, os_event) ~ 1, data = dat_lasso)

    bic <- -2 * as.numeric(logLik(fit_null)) + log(n) * k

    return(list(
      penalty = pen,
      fit = fit.bar,
      bic = bic,
      refit_warn = FALSE,
      refit_ok = TRUE
    ))
  } else {

    dat_bar <- data.frame(
      os_months = dat_lasso$os_months,
      os_event = dat_lasso$os_event,
      X[, selected_idx, drop = FALSE]
    )

    colnames(dat_bar)[3:ncol(dat_bar)] <- colnames(X)[selected_idx]

    warn_msg <- NULL

    fit_refit <- tryCatch(
      withCallingHandlers(
        coxph(
          as.formula(
            paste(
              "Surv(os_months, os_event) ~",
              paste(colnames(X)[selected_idx], collapse = " + ")
            )
          ),
          data = dat_bar
        ),
        warning = function(w) {
          warn_msg <<- conditionMessage(w)
          invokeRestart("muffleWarning")
        }
      ),
      error = function(e) NULL
    )

    if(is.null(fit_refit)) {
      bic <- Inf
      refit_ok <- FALSE
    } else {
      bic <- -2 * as.numeric(logLik(fit_refit)) + log(n) * k
      refit_ok <- TRUE
    }

    return(list(
      penalty = pen,
      fit = fit.bar,
      bic = bic,
      refit_warn = !is.null(warn_msg),
      warn_msg = warn_msg,
      refit_ok = refit_ok
    ))
  }
})
```

```{r}
bic_values <- sapply(coxbar_grid, function(x) x$bic)

best_index <- which.min(bic_values)

pen_grid[best_index]
range(pen_grid)

best_index
length(pen_grid)
```

best fit model:
```{r}
fit.coxbar <- coxbar_grid[[best_index]]$fit

coef_bar <- coef(fit.coxbar)
```

selected variables:
```{r}
coxbar_selected <- colnames(X)[coef_bar != 0]

coxbar_selected

coxbar_selected_df <- data.frame(
  variable = colnames(X)[coef_bar != 0],
  coef = coef_bar[coef_bar != 0]
) %>%
  arrange(desc(abs(coef)))

coxbar_selected_df
```

```{r}
sum(is.infinite(bic_values))

sum(sapply(coxbar_grid, function(x) x$refit_warn), na.rm = TRUE)

sum(sapply(coxbar_grid, function(x) x$refit_ok), na.rm = TRUE)
```

Only 2 of the 50 penalty values failed during fitting.
Only 1 run produced an infinite-coefficient warning in the refitted Cox model.
Overall, 48 penalty values produced stable refitted Cox models.

### compare methods
```{r}
gene_sets <- list(
  lasso = lasso_selected,
  scad = scad_selected,
  mcp = mcp_selected,
  coxbar = coxbar_selected
)
```

```{r}
refit_compare <- lapply(names(gene_sets), function(m) {

  vars <- gene_sets[[m]]

  dat_tmp <- dat_lasso %>%
    select(os_months, os_event, all_of(vars))

  fit_tmp <- coxph(
    as.formula(
      paste("Surv(os_months, os_event) ~", paste(vars, collapse = " + "))
    ),
    data = dat_tmp
  )

  data.frame(
    method = m,
    n_var = length(vars),
    logLik = as.numeric(logLik(fit_tmp)),
    AIC = AIC(fit_tmp),
    BIC = BIC(fit_tmp)
  )
}) %>%
  bind_rows() %>%
  arrange(BIC)

refit_compare
```


### select variables
Our aim was to select 5 to 8 final gene variables for inclusion in the Cox model.


why choose CoxBAR as the main model:
In addition that coxBAR gives best BIC, the main reason is because our goal was sparse and stable variable screening in a correlated genomic setting, we treated CoxBAR as the primary penalized Cox selection method. CoxBAR is a surrogate ℓ0 penalized Cox approach based on broken adaptive ridge and has been shown to have selection consistency, oracle property for parameter estimation, and a grouping property for highly correlated covariates. In simulation and empirical evaluations, BAR-based Cox regression selected relatively sparse models and achieved model selection and discriminatory performance comparable to or slightly better than Cox LASSO, with less biased and more stable solution paths under BIC-based tuning.

summarize overlap among selected variables:
```{r}
all_selected <- list(
  lasso = lasso_selected,
  scad = scad_selected,
  mcp = mcp_selected,
  coxbar = coxbar_selected
)

all_genes <- sort(unique(unlist(all_selected)))

overlap_df <- data.frame(variable = all_genes) %>%
  mutate(
    in_lasso = variable %in% lasso_selected,
    in_scad = variable %in% scad_selected,
    in_mcp = variable %in% mcp_selected,
    in_coxbar = variable %in% coxbar_selected,
    n_methods = in_lasso + in_scad + in_mcp + in_coxbar
  ) %>%
  arrange(desc(n_methods), variable)

overlap_df
```


The variables selected by all 4 methods were tsc2, bbc3, sbno1, ptprd_mut, and nek1_mut.


The final selected variables were tsc2, bbc3, sbno1, ptprd_mut, nek1_mut. This set was primarily based on CoxBAR and also showed good stability across selection methods, since the first five variables appeared in all four methods and the remaining two appeared in three methods.

Because ptprd_mut and nek1_mut were originally recorded as either 0 or mutation-type strings, they were recoded as binary mutation indicators, with 1 indicating mutation and 0 indicating no mutation, before being entered into the Cox model.

## Crude Cox with selected genes
### set up data
```{r}
hx_dat0 <- METABRIC_RNA_Mutation %>%
  as_tibble()

hx_dat1 <- hx_dat0 %>%
  transmute(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event  = overall_survival,

    age = age_at_diagnosis,
    tumor_size,
    nodes_positive = lymph_nodes_examined_positive,
    grade = neoplasm_histologic_grade,
    her2 = her2_status,
    pam50 = `pam50_._claudin.low_subtype`,
    tumor_stage,
    menopausal_state = inferred_menopausal_state,
    pr_status,
    surgery = type_of_breast_surgery,
    radio_therapy,

    tsc2,
    bbc3,
    sbno1,
    ptprd_mut,
    nek1_mut,
    mapk14,
    kmt2c
  )

hx_dat2 <- hx_dat1 %>%
  mutate(
    across(where(is.character), ~ na_if(.x, "")),

    er_status       = as.character(er_status),
    hormone_therapy = as.character(hormone_therapy),
    chemotherapy    = as.character(chemotherapy),

    grade            = as.character(grade),
    her2             = as.character(her2),
    pam50            = as.character(pam50),
    tumor_stage      = as.character(tumor_stage),
    menopausal_state = as.character(menopausal_state),
    pr_status        = as.character(pr_status),
    surgery          = as.character(surgery),
    radio_therapy    = as.character(radio_therapy),

    os_months      = as.numeric(os_months),
    os_event       = as.numeric(os_event),
    age            = as.numeric(age),
    tumor_size     = as.numeric(tumor_size),
    nodes_positive = as.numeric(nodes_positive),

    tsc2   = as.numeric(tsc2),
    bbc3   = as.numeric(bbc3),
    sbno1  = as.numeric(sbno1),
    mapk14 = as.numeric(mapk14),
    kmt2c  = as.numeric(kmt2c)
  ) %>%
  mutate(
    hormone_therapy_num = case_when(
      hormone_therapy %in% c("1", 1) ~ 1,
      hormone_therapy %in% c("0", 0) ~ 0,
      TRUE ~ NA_real_
    ),
    chemotherapy_num = case_when(
      chemotherapy %in% c("1", 1) ~ 1,
      chemotherapy %in% c("0", 0) ~ 0,
      TRUE ~ NA_real_
    ),
    ptprd_mut_bin = case_when(
      is.na(ptprd_mut) ~ NA_real_,
      ptprd_mut %in% c("0", 0) ~ 0,
      TRUE ~ 1
    ),
    nek1_mut_bin = case_when(
      is.na(nek1_mut) ~ NA_real_,
      nek1_mut %in% c("0", 0) ~ 0,
      TRUE ~ 1
    )
  )

hx_dat_er <- hx_dat2 %>%
  filter(
    er_status == "Positive",
    hormone_therapy_num == 1
  )
```


```{r}
# set up data for crude cox and aft
hx_dat_model <- hx_dat_er %>%
  filter(
    !is.na(os_months),
    !is.na(os_event),
    !is.na(chemotherapy_num),
    !is.na(age),
    !is.na(tumor_size),
    !is.na(nodes_positive),
    !is.na(grade),
    !is.na(pam50),
    !is.na(tsc2),
    !is.na(bbc3),
    !is.na(sbno1),
    !is.na(ptprd_mut_bin),
    !is.na(nek1_mut_bin)
  ) %>%
  mutate(
    chemo = factor(
      chemotherapy_num,
      levels = c(0, 1),
      labels = c("Hormone only", "Hormone + Chemo")
    ),
    chemo_num = chemotherapy_num,
    grade = factor(grade),
    pam50_grp = case_when(
      pam50 %in% c("LumA", "LumB") ~ pam50,
      TRUE ~ "Other"
    )
  ) %>%
  mutate(
    pam50_grp = factor(pam50_grp, levels = c("LumA", "LumB", "Other")),
    ptprd_mut_bin = factor(ptprd_mut_bin, levels = c(0, 1), labels = c("No", "Yes")),
    nek1_mut_bin  = factor(nek1_mut_bin,  levels = c(0, 1), labels = c("No", "Yes"))
  ) %>%
  filter(!is.na(pam50_grp))

nrow(hx_dat_model)
table(hx_dat_model$chemo, useNA = "ifany")
table(hx_dat_model$grade, useNA = "ifany")
table(hx_dat_model$pam50_grp, useNA = "ifany")
```


### Cox model without time-dependent chemotherapy
```{r}
# fit crude cox model
hx_cox_formula <- Surv(os_months, os_event) ~
  chemo +
  age +
  tumor_size +
  nodes_positive +
  grade +
  pam50_grp +
  tsc2 +
  bbc3 +
  sbno1 +
  ptprd_mut_bin +
  nek1_mut_bin

hx_cox_fit <- coxph(
  hx_cox_formula,
  data = hx_dat_model,
  ties = "breslow"
)

summary(hx_cox_fit)
```

```{r}
# vif check
hx_lm_proxy <- lm(
  os_months ~ chemo + age + tumor_size + nodes_positive + grade +
    pam50_grp + tsc2 + bbc3 + sbno1 + ptprd_mut_bin + nek1_mut_bin,
  data = hx_dat_model
)

vif(hx_lm_proxy)
```

```{r}
# ph assumption check
hx_ph_res <- cox.zph(hx_cox_fit)
hx_ph_res
plot(hx_ph_res)
```



### Cox model with time-dependent chemotherapy
```{r}
# time-dependent cox
hx_event_times <- sort(unique(hx_dat_model$os_months[hx_dat_model$os_event == 1]))
hx_event_times <- hx_event_times[-1]

hx_dat_split <- survSplit(
  data  = hx_dat_model,
  cut   = hx_event_times,
  end   = "os_months",
  start = "t0",
  event = "os_event"
) %>%
  mutate(
    chemo_num = ifelse(chemo == "Hormone + Chemo", 1, 0),
    chemo_t   = chemo_num * log(os_months)
  )
```

```{r}
hx_cox_td_fit <- coxph(
  Surv(t0, os_months, os_event) ~
    chemo_num +
    chemo_t +
    age +
    tumor_size +
    nodes_positive +
    grade +
    pam50_grp +
    tsc2 +
    bbc3 +
    sbno1 +
    ptprd_mut_bin +
    nek1_mut_bin,
  data = hx_dat_split,
  ties = "breslow"
)

summary(hx_cox_td_fit)
```

```{r}
# ph check for td cox
hx_ph_td_res <- cox.zph(hx_cox_td_fit)
hx_ph_td_res
plot(hx_ph_td_res)
```

### Anderson's plot
```{r}
# draw anderson plot
hx_reptime <- function(l, t, range_of_time) {
  x <- numeric(length(range_of_time))
  for (i in seq_along(range_of_time)) {
    dif <- range_of_time[i] - t
    dif <- dif[dif >= 0]
    if (length(dif) > 0) {
      x[i] <- l[which.min(dif)]
    } else {
      x[i] <- 0
    }
  }
  x
}

hx_dat_anderson <- hx_dat_model %>%
  mutate(chemo_num = ifelse(chemo == "Hormone + Chemo", 1, 0))

hx_fit_and <- basehaz(
  coxph(
    Surv(os_months, os_event) ~ strata(chemo_num),
    data = hx_dat_anderson,
    ties = "breslow"
  ),
  centered = FALSE
)

hx_H0 <- hx_fit_and$hazard[hx_fit_and$strata == "chemo_num=0"]
hx_H1 <- hx_fit_and$hazard[hx_fit_and$strata == "chemo_num=1"]

hx_t0 <- hx_fit_and$time[hx_fit_and$strata == "chemo_num=0"]
hx_t1 <- hx_fit_and$time[hx_fit_and$strata == "chemo_num=1"]

hx_grid_time <- sort(unique(c(hx_t0, hx_t1)))

hx_H0_rep <- hx_reptime(hx_H0, hx_t0, hx_grid_time)
hx_H1_rep <- hx_reptime(hx_H1, hx_t1, hx_grid_time)

plot(
  hx_H1_rep ~ hx_H0_rep,
  main = "Anderson Plot for Chemotherapy Groups",
  xlab = "Cumulative Hazard: Hormone only",
  ylab = "Cumulative Hazard: Hormone + Chemo",
  type = "s"
)
abline(0, 1, lty = 2, col = "red")
```

### Cox-snell residual plot
```{r}
# draw cox-snell residual plot
hx_cs_resid <- predict(hx_cox_td_fit, type = "expected")

hx_fit_cs <- survfit(Surv(hx_cs_resid, os_event) ~ 1, data = hx_dat_split)

hx_H_cs <- cumsum(hx_fit_cs$n.event / hx_fit_cs$n.risk)

plot(
  hx_fit_cs$time, hx_H_cs,
  type = "s",
  col = "blue",
  main = "Cox-Snell Residual Plot\nFinal Time-Dependent Cox Model",
  xlab = "Cox-Snell Residual",
  ylab = "Nelson-Aalen Cumulative Hazard"
)
abline(0, 1, col = "red", lty = 2)
```

## AFT models
```{r}
# fit aft models
hx_aft_formula <- Surv(os_months, os_event) ~
  chemo +
  age +
  tumor_size +
  nodes_positive +
  grade +
  pam50_grp +
  tsc2 +
  bbc3 +
  sbno1 +
  ptprd_mut_bin +
  nek1_mut_bin

hx_fit_exp <- survreg(hx_aft_formula, data = hx_dat_model, dist = "exponential")
hx_fit_weib <- survreg(hx_aft_formula, data = hx_dat_model, dist = "weibull")
hx_fit_lnorm <- survreg(hx_aft_formula, data = hx_dat_model, dist = "lognormal")
hx_fit_llogis <- survreg(hx_aft_formula, data = hx_dat_model, dist = "loglogistic")
```

```{r}
# compare aft models
hx_model_compare <- data.frame(
  Model = c("Exponential", "Weibull", "Log-normal", "Log-logistic"),
  AIC = c(AIC(hx_fit_exp), AIC(hx_fit_weib), AIC(hx_fit_lnorm), AIC(hx_fit_llogis)),
  BIC = c(BIC(hx_fit_exp), BIC(hx_fit_weib), BIC(hx_fit_lnorm), BIC(hx_fit_llogis)),
  logLik = c(logLik(hx_fit_exp), logLik(hx_fit_weib), logLik(hx_fit_lnorm), logLik(hx_fit_llogis))
) %>%
  arrange(AIC)

hx_model_compare
```

```{r}
# summarize aft models
summary(hx_fit_exp)
summary(hx_fit_weib)
summary(hx_fit_lnorm)
summary(hx_fit_llogis)
```

```{r}
summary(hx_fit_weib)
```


```{r}
# extract aft time ratios
hx_extract_aft_table <- function(fit) {
  tb <- as.data.frame(summary(fit)$table)
  tb$term <- rownames(tb)

  tb %>%
    mutate(
      lower = Value - 1.96 * `Std. Error`,
      upper = Value + 1.96 * `Std. Error`,
      TR = exp(Value),
      TR_lower = exp(lower),
      TR_upper = exp(upper)
    ) %>%
    select(term, Value, `Std. Error`, p, TR, TR_lower, TR_upper)
}

hx_aft_exp_table <- hx_extract_aft_table(hx_fit_exp)
hx_aft_weib_table <- hx_extract_aft_table(hx_fit_weib)
hx_aft_lnorm_table <- hx_extract_aft_table(hx_fit_lnorm)
hx_aft_llogis_table <- hx_extract_aft_table(hx_fit_llogis)

hx_aft_weib_table
```


```{r}
# plot adjusted survival from weibull aft

library(flexsurv)

hx_fit_weib_flex <- flexsurvreg(
  hx_aft_formula,
  data = hx_dat_model,
  dist = "weibull"
)

hx_newdata_typical <- data.frame(
  chemo = factor(c("Hormone only", "Hormone + Chemo"),
                 levels = levels(hx_dat_model$chemo)),
  age = mean(hx_dat_model$age, na.rm = TRUE),
  tumor_size = mean(hx_dat_model$tumor_size, na.rm = TRUE),
  nodes_positive = mean(hx_dat_model$nodes_positive, na.rm = TRUE),
  grade = factor(levels(hx_dat_model$grade)[1], levels = levels(hx_dat_model$grade)),
  pam50_grp = factor("LumA", levels = levels(hx_dat_model$pam50_grp)),
  tsc2 = mean(hx_dat_model$tsc2, na.rm = TRUE),
  bbc3 = mean(hx_dat_model$bbc3, na.rm = TRUE),
  sbno1 = mean(hx_dat_model$sbno1, na.rm = TRUE),
  ptprd_mut_bin = factor("No", levels = levels(hx_dat_model$ptprd_mut_bin)),
  nek1_mut_bin = factor("No", levels = levels(hx_dat_model$nek1_mut_bin))
)

hx_pred <- summary(
  hx_fit_weib_flex,
  newdata = hx_newdata_typical,
  type = "survival",
  t = seq(0, 250, by = 1)
)

plot(
  hx_pred[[1]]$time, hx_pred[[1]]$est,
  type = "l", lwd = 2,
  xlab = "Time since diagnosis (months)",
  ylab = "Adjusted survival probability",
  main = "Adjusted Survival Curves from Weibull AFT"
)

lines(hx_pred[[2]]$time, hx_pred[[2]]$est, lwd = 2, lty = 2)

legend(
  "topright",
  legend = c("Hormone only", "Hormone + Chemo"),
  lwd = 2,
  lty = c(1, 2)
)
```


## Model Comparisons
### c-index
```{r}
# clinical-only time-dependent cox
ccox_td_score <- -predict(ccox_td_fit, type = "lp")
ccox_td_c <- concordance(
  Surv(t0, os_months, os_event) ~ ccox_td_score,
  data = dat_ccox_split
)

# weighted time-dependent cox
cox_w_td_score <- -predict(cox_w_td, type = "lp")
cox_w_td_c <- concordance(
  Surv(t0, os_months, os_event) ~ cox_w_td_score,
  data = dat_split_w
)

# crude cox + gene time-dependent
hx_cox_td_score <- -predict(hx_cox_td_fit, type = "lp")
hx_cox_td_c <- concordance(
  Surv(t0, os_months, os_event) ~ hx_cox_td_score,
  data = hx_dat_split
)

# aft (weibull)
weib_score <- predict(hx_fit_weib, type = "lp")
weib_c <- concordance(
  Surv(os_months, os_event) ~ weib_score,
  data = hx_dat_model
)

# combine results
cindex_compare <- data.frame(
  model = c(
    "Clinical-only TD Cox",
    "Weighted TD Cox",
    "TD Cox + Gene",
    "AFT Weibull"
  ),
  c_index = c(
    ccox_td_c$concordance,
    cox_w_td_c$concordance,
    hx_cox_td_c$concordance,
    weib_c$concordance
  ),
  se = c(
    sqrt(ccox_td_c$var),
    sqrt(cox_w_td_c$var),
    sqrt(hx_cox_td_c$var),
    sqrt(weib_c$var)
  )
)

cindex_compare
```

The concordance indices indicate moderate discriminative performance across the survival models. The clinical-only time-dependent Cox model achieved a C-index of 0.613, while the IPTW weighted time-dependent Cox model produced a very similar value of 0.607, suggesting that weighting to address confounding by indication had little impact on the model’s overall predictive discrimination. In contrast, incorporating the selected gene variables substantially improved model performance. The time-dependent Cox model with gene predictors achieved a higher C-index of 0.676, indicating better ability to distinguish patients with different survival outcomes. The Weibull AFT model showed comparable discrimination with a C-index of 0.671, suggesting that the parametric AFT framework provides similar predictive capability to the Cox model when using the same set of predictors. Overall, these results suggest that while propensity score weighting primarily adjusts treatment effect estimation, the inclusion of selected gene variables meaningfully improves the model’s ability to predict survival.






