---
title: "Project part"
author: "Hanxi Chen"
subtitle: :)
date: today
format:
  pdf:
    documentclass: article
    number-sections: false
    fig-align: center
    keep-tex: false  # keeps the intermediate .tex file
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    html-math-method: mathjax
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r, warning=FALSE}
library(tidyverse)
library(car)
library(dplyr)
library(survival)
library(survminer)
library(zoo) # na.locf

library(janitor)
library(skimr)

library(tableone)

library(tableone)
library(gtsummary)
```


## Import and Cleaning

```{r}
METABRIC_RNA_Mutation <- read.csv("METABRIC_RNA_Mutation.csv")
```

```{r}
dat0 <- METABRIC_RNA_Mutation %>%
  as_tibble()

# rename
dat1 <- dat0 %>%
  transmute(
    patient_id,
    er_status,
    hormone_therapy,
    chemotherapy,
    os_months = overall_survival_months,
    os_event = overall_survival,
    
    age = age_at_diagnosis,
    tumor_size,
    nodes_positive = lymph_nodes_examined_positive,
    grade = neoplasm_histologic_grade,
    her2 = her2_status,
    pam50 = `pam50_._claudin.low_subtype`,
    tumor_stage,
    menopausal_state = inferred_menopausal_state,
    pr_status,
    surgery = type_of_breast_surgery,
    radio_therapy
  )

# cleaning
dat2 <- dat1 %>%
  mutate(
    across(where(is.character), ~na_if(.x, "")),
    across(c(er_status, hormone_therapy, chemotherapy,
             grade, her2, pam50, tumor_stage,
             menopausal_state, pr_status, surgery, radio_therapy), as.character),
    across(c(age, tumor_size, nodes_positive, os_months, os_event), as.numeric)
  )
```

```{r}
table(dat2$er_status, useNA = "ifany")
table(dat2$hormone_therapy, useNA = "ifany")
table(dat2$chemotherapy, useNA = "ifany")
table(dat2$grade, useNA = "ifany")
table(dat2$her2, useNA = "ifany")
table(dat2$pam50, useNA = "ifany")
table(dat2$tumor_stage, useNA = "ifany")
table(dat2$menopausal_state, useNA = "ifany")
table(dat2$pr_status, useNA = "ifany")
table(dat2$surgery, useNA = "ifany")
table(dat2$radio_therapy, useNA = "ifany")
```
```{r}
dat_table1 <- dat2 %>%
  filter(
    er_status == "Positive",
    hormone_therapy == 1
  ) %>%
  mutate(
    chemo = chemotherapy,
    trt_group = factor(
      chemo,
      levels = c(0, 1),
      labels = c("Hormone only", "Hormone + Chemo")
    )
  ) %>%
  filter(
    !is.na(chemo),
    !is.na(os_months),
    !is.na(os_event)
  )

nrow(dat0)
nrow(dat_table1)

table(dat_table1$trt_group, useNA = "ifany")
prop.table(table(dat_table1$trt_group))
```
```{r}
vars_check <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status",
  "surgery", "radio_therapy"
)

missing_by_group_long <- dat_table1 %>%
  group_by(trt_group) %>%
  summarise(
    across(all_of(vars_check), ~mean(is.na(.))*100)
  ) %>%
  pivot_longer(
    -trt_group,
    names_to = "variable",
    values_to = "pct_missing"
  )

missing_by_group_long
```

## Table 1: covariate distributions over Trt Groups

```{r}
# 废案
vars <- c(
  "age", "tumor_size", "nodes_positive",
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

factorVars <- c(
  "grade", "her2", "pam50", "tumor_stage",
  "menopausal_state", "pr_status", "surgery", "radio_therapy"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "trt_group",
  data = dat_table1,
  factorVars = factorVars,
  includeNA = TRUE
)

print(
  tab1,
  showAllLevels = TRUE,
  quote = FALSE,
  noSpaces = TRUE,
  nonnormal = c("tumor_size", "nodes_positive")
)
```

##### numeric table
Patients receiving chemotherapy were significantly younger (mean age 53.6 vs 66.1 years, p < 0.001) and had more aggressive tumor features, including larger tumor size, greater lymph node involvement, and higher histologic grade (all p < 0.001). Pre-menopausal status was also more common among patients receiving chemotherapy (38.2% vs 8.0%, p < 0.001). HER2 status and PR status did not significantly differ between treatment groups.

These findings suggest that treatment assignment was strongly associated with baseline risk factors.

```{r}
summary_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    g1 <- as.numeric(x[g == "Hormone only"])
    g2 <- as.numeric(x[g == "Hormone + Chemo"])
    
    ttest <- t.test(g1, g2)
    
    mean1 <- mean(g1, na.rm = TRUE)
    sd1 <- sd(g1, na.rm = TRUE)
    mean2 <- mean(g2, na.rm = TRUE)
    sd2 <- sd(g2, na.rm = TRUE)
    smd <- (mean2 - mean1) / sqrt((sd1^2 + sd2^2) / 2)
    
    data.frame(
      Variable = v,
      Mean_Group1 = mean1,
      SD_Group1 = sd1,
      Mean_Group2 = mean2,
      SD_Group2 = sd2,
      t_value = as.numeric(ttest$statistic),
      p_value = ttest$p.value,
      SMD = smd
    )
  })
  
  dplyr::bind_rows(out)
}

cont_vars <- c("age", "tumor_size", "nodes_positive")

cont_table <- summary_table(dat_table1, cont_vars) |>
  dplyr::mutate(
    across(where(is.numeric), ~round(.x, 3))
  )

print(cont_table)
```


##### categorical table
```{r}
categorical_table <- function(data, vars, group_var = "trt_group") {
  
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    g <- data[[group_var]]
    
    # 变成字符，保留 NA 为单独一类
    x_chr <- as.character(x)
    x_chr[is.na(x_chr)] <- "Missing"
    
    g_chr <- as.character(g)
    
    tab <- table(x_chr, g_chr)
    prop <- prop.table(tab, margin = 2) * 100
    
    # 自动选择检验
    test <- if (any(tab < 5)) fisher.test(tab) else chisq.test(tab)
    
    stat_value <- if ("statistic" %in% names(test)) as.numeric(test$statistic) else NA_real_
    p_value <- test$p.value
    
    # 防止列名顺序或缺失导致报错
    n1 <- if ("Hormone only" %in% colnames(tab)) tab[, "Hormone only"] else rep(0, nrow(tab))
    n2 <- if ("Hormone + Chemo" %in% colnames(tab)) tab[, "Hormone + Chemo"] else rep(0, nrow(tab))
    
    p1 <- if ("Hormone only" %in% colnames(prop)) prop[, "Hormone only"] else rep(0, nrow(prop))
    p2 <- if ("Hormone + Chemo" %in% colnames(prop)) prop[, "Hormone + Chemo"] else rep(0, nrow(prop))
    
    df <- data.frame(
      Variable = v,
      Level = rownames(tab),
      n_Group1 = as.numeric(n1),
      pct_Group1 = as.numeric(p1),
      n_Group2 = as.numeric(n2),
      pct_Group2 = as.numeric(p2),
      statistic = NA_real_,
      p_value = NA_real_,
      stringsAsFactors = FALSE
    )
    
    # 只在每个变量第一行放检验结果
    df$statistic[1] <- stat_value
    df$p_value[1] <- p_value
    
    df
  })
  
  dplyr::bind_rows(out)
}

cat_vars <- c(
  "grade",
  "her2",
  "pam50",
  "tumor_stage",
  "menopausal_state",
  "pr_status",
  "surgery",
  "radio_therapy"
)

cat_table <- categorical_table(dat_table1, cat_vars) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~round(.x, 3))
  )

print(cat_table)
```




### gene variable selection: LASSO/SCAD/MCP/coxBAR



